Readers already familiar with GHC's concurrency primitives may find it
enough to skim this section noting the syntactic differences in the
\dejafu{} variant.

\begin{departure}
  The few departures from the semantics of the traditional concurrency
  abstraction are highlighted like this.
\end{departure}

The \verb|MonadConc| typeclass has an instance\footnote{A typeclass
  has \emph{instances}, each type may have one unique instance for a
  typeclass.} for \verb|IO|, and so existing code using only the
functions generalised over can be made suitable for testing quite
simply. Existing code which makes use of more functionality may
require a light dusting of \verb|liftIO|\footnote{The \texttt{IO} type
  allows unrestricted side-effects during execution. It turns out that
  many useful types are just \texttt{IO} with some extra structure
  applied, and the \texttt{liftIO} function (which belongs to a
  typeclass called \texttt{MonadIO}) can be used to `translate' the
  effects into such a type.} where it is safe to do so, which will be
expanded upon in \sect{abstraction}{typeclass-lifting}.

\subsubsection{Threads}
\label{sec:abstraction-typeclass-threads}

Threads let a program do multiple things at once. Every program has at
least one thread, which starts where \verb|main| does and runs until
the program terminates. A thread is the basic unit of concurrency. It
lets us pretend (with parallelism, it might even be true!) that we're
computing multiple things at once.

We can start a new thread with the function:\footnote{This is a
  function named \texttt{fork} with a \emph{type signature}. Type
  signatures may contain typeclass constraints, type variables, type
  constructors (similar to generics in other languages), and concrete
  types. Here \texttt{ThreadId m} is a type constructor applied to the
  type variable \texttt{m}, which is constrained to be a type with an
  instance of \texttt{MonadConc}.}

\begin{haskellcode}
fork :: MonadConc m => m () -> m (ThreadId m)
\end{haskellcode}

This starts evaluating its argument in a separate thread. It also
gives us back a (monad-specific) \verb|ThreadId| value, which we can
use to kill the thread later on, if we want.

A thread can query its own \verb|ThreadId|:

\begin{haskellcode}
myThreadId :: MonadConc m => m (ThreadId m)
\end{haskellcode}

In a real machine, there are of course a number of processors and
cores. It may be that a particular application of concurrency is only
a net gain if every thread is operating on a separate core, so that
threads are not interrupting each other. The GHC runtime refers to the
number of Haskell threads that can run truly simultaneously as the
number of \emph{capabilities}. We can query this value, and fork
threads which are bound to a particular capability:

\begin{haskellcode}
getNumCapabilities :: MonadConc m => m Int
forkOn :: MonadConc m => Int -> m () -> m (ThreadId m)
\end{haskellcode}

% Layout hack - line is too long otherwise
The \verb|forkOn| function interprets the capability number modulo the
value returned by \\\verb|getNumCapabilities|.

\begin{departure}
  \verb|getNumCapabilities| is not required to return a true
  result. The testing instances return ``2'' despite executing
  everything in the same capability, to encourage more
  concurrency. The \verb|IO| instance does return a true result.
\end{departure}

Sometimes we just want the special case of evaluating something in a
separate thread, for which we can use \verb|spawn| (implemented in
terms of \verb|fork|):

\begin{haskellcode}
spawn :: MonadConc m => m a -> m (CVar m a)
\end{haskellcode}

This returns a \verb|CVar| (\emph{Concurrent Variable}), to which we
can apply \verb|readCVar|, blocking until the computation is done and
the value is stored.

Threads are scheduled non-deterministically, where every time the
runtime decides to perform a context switch, one of the runnable
threads will be executed. Sometimes, however, a thread may be runnable
but also waiting for something to happen. The programmer can provide a
clue to the scheduler that another thread should be tried instead:

\begin{haskellcode}
yield :: MonadConc m => m ()
\end{haskellcode}

This gives any other thread the opportunity to execute instead of the
yielding one, but it is not \emph{required} to cause a context switch
except on co-operative multitasking systems.

\subsubsection{Threading and the Foreign Function Interface}
\label{sec:abstraction-typeclass-ffithreads}

In order to accommodate Foreign Function Interface (FFI) calls which
may block, GHC provides a mechanism for \emph{binding} a Haskell
thread to an operating system thread. This allows FFI calls to be
managed by the operating system, unlike normal Haskell threads which
are managed by the runtime and multiplexed onto a smaller number of
operating system threads. This means that blocking FFI calls do not
necessarily block the entire program.

There is no \verb|MonadConc| equivalent of bound threads, as there
would be no way to reliably test this behaviour. Unfortunately, if
bound threads are required, \verb|IO| will have to be used.

A few predicates are provided for compatibility:\footnote{Function
  application in Haskell uses no special syntax, only juxtaposition,
  so \texttt{return False} is applying the value \texttt{False} to the
  function \texttt{return}. \texttt{return} here is used to inject a
  value into a monad, it is unfortunately named and has nothing to do
  with the return keyword in other languages.}

\begin{haskellcode}
rtsSupportsBoundThreads :: Bool
rtsSupportsBoundThreads = False

isCurrentThreadBound :: MonadConc m => m Bool
isCurrentThreadBound = return False
\end{haskellcode}

\subsubsection{Mutable State}
\label{sec:abstraction-typeclass-crefs}

Threading by itself is not really enough. We need to be able to
\emph{communicate} between threads: we've already seen an instance
of this with the \verb|spawn| function.

The simplest type of mutable shared state provided is the \verb|CRef|
(\emph{Concurrent Reference}). \verb|CRef|s are shared variables which
can be written to and read from:

\begin{haskellcode}
newCRef    :: MonadConc m => a -> m (CRef m a)
readCRef   :: MonadConc m => CRef m a -> m a
modifyCRef :: MonadConc m => CRef m a -> (a -> (a, b)) -> m b
writeCRef  :: MonadConc m => CRef m a -> a -> m ()
\end{haskellcode}

The \verb|modifyCRef| function is atomic. The \verb|readCRef| and
\verb|writeCRef| functions are not synchronised: it is possible for
one thread to read from a \verb|CRef| strictly after another thread
has written to it and to observe an old value!  This is expanded more
upon in \sect{abstraction}{mem}. To ensure that every thread sees a
value as soon as it is written there is a synchronised write function:

\begin{haskellcode}
atomicWriteCRef :: MonadConc m => CRef m a -> a -> m ()
\end{haskellcode}

However, synchronisation can slow down execution in a parallel
environment. Note that \verb|modifyCRef| is also synchronised.

\subsubsection{Compare-and-swap (CAS)}
\label{sec:abstraction-typeclass-cas}

As \verb|CRef|s correspond very closely to mutable memory locations,
there is also a compare-and-swap interface available. Compare-and-swap
is a synchronised atomic primitive which is used to update a location
in memory if and only if it has not been changed since some witness
value was produced. This role of this witness value is called a
\verb|Ticket| here:

\begin{haskellcode}
readForCAS :: MonadConc m => CRef m a -> m (Ticket m a)
peekTicket :: MonadConc m => Ticket m a -> m a
\end{haskellcode}

A \verb|Ticket| can be used to check if a \verb|CRef| has been written
to since it was produced, and can also be used to get the value that
was seen then.

\begin{haskellcode}
casCRef :: MonadConc m => CRef m a -> Ticket m a -> a -> m (Bool, Ticket m a)
\end{haskellcode}

The \verb|casCRef| function is synchronised, is strict in the value
written, and will replace the value within a \verb|CRef| if it hasn't
been modified since the \verb|Ticket| was produced. It returns an
indication of success and a \verb|Ticket| to use in future
operations. This operation is often used in the implementation of
lock-free synchronisation primitives.

There is also an equivalent of \verb|modifyCRef| using a
compare-and-swap. This behaves almost the same as the non-CAS version
but may be more performant in some cases, and is strict in the value
being written:

\begin{haskellcode}
modifyCRefCAS :: MonadConc m => CRef m a -> (a -> (a, b)) -> m b
\end{haskellcode}

\subsubsection{Mutual Exclusion}
\label{sec:abstraction-typeclass-cvars}

A \verb|CVar| is a shared variable under \emph{mutual exclusion}. It
has two possible states: \emph{full} or \emph{empty}. Writing to a
full \verb|CVar| blocks until it is empty, and reading or taking from
an empty \verb|CVar| blocks until it is full. There are also
non-blocking functions which return an indication of success:

\begin{haskellcode}
newEmptyCVar :: MonadConc m => m (CVar m a)
putCVar      :: MonadConc m => CVar m a -> a -> m ()
readCVar     :: MonadConc m => CVar m a -> m a
takeCVar     :: MonadConc m => CVar m a -> m a
tryPutCVar   :: MonadConc m => CVar m a -> a -> m Bool
tryTakeCVar  :: MonadConc m => CVar m a -> m (Maybe a)
\end{haskellcode}

Unfortunately, the mutual exclusion behaviour of \verb|CVar|s means
that computations can become \emph{deadlocked}. For example, deadlock
occurs if every thread tries to take from the same \verb|CVar|. The
GHC runtime can detect this in some situations (and will complain if
it does), and so can \dejafu{} in a more informative way.

\begin{departure}
  \dejafu{} can only detect deadlock to the same extent as GHC if
  every thread is annotated with which \verb|CVar|s it knows
  about. This is because GHC uses the garbage collector to solve this
  problem, which is out of the reach of \dejafu{}.
\end{departure}

There are also additional functions provided in the
Control\-.Concurrent\-.CVar and Control\-.Concurrent\-.CVar\-.Strict
modules.

\subsubsection{Exceptions}
\label{sec:abstraction-typeclass-excs}

Exceptions are a way to bail out of a computation early. Whether
they're a good solution to that problem is a question of style, but
they can be used to jump quickly to error handling code when
necessary. The basic functions for dealing with exceptions are:

\begin{haskellcode}
catch :: (Exception e, MonadConc m) => m a -> (e -> m a) -> m a
throw :: (Exception e, MonadConc m) => e -> m a
\end{haskellcode}

Where \verb|throw| causes the computation to jump back to the nearest
enclosing \verb|catch| capable of handling the particular
exception. As exceptions belong to a typeclass, rather than being a
concrete type, different \verb|catch| functions can be nested, to
handle different types of exceptions.

\begin{departure}
  The \verb|IO| \verb|catch| function can catch exceptions from pure
  code. This is not true in general for \verb|MonadConc| instances.
  So some things which work normally may not work in testing, and we
  risk false negatives. This is a small cost, however, as exceptions
  from pure code are things like pattern match failures and evaluating
  \verb|undefined|, which are arguably bugs.
\end{departure}

Exceptions can be used to kill a thread:

\begin{haskellcode}
throwTo :: (Exception e, MonadConc m) => ThreadId m -> e -> m ()
killThread :: MonadConc m => ThreadId m -> m ()
\end{haskellcode}

These functions block until the target thread is in an appropriate
state to receive the exception.

What if we don't want our threads to be subject to destruction in this
way? A thread also has a \emph{masking state}, which can be used to
block exceptions from other threads. There are three masking states:
\emph{unmasked}, in which a thread can have exceptions thrown to it;
\emph{interruptible}, in which a thread can only have exceptions
thrown to it if it is blocked; and \emph{uninterruptible}, in which a
thread cannot have exceptions thrown to it. When a thread is started,
it inherits the masking state of its parent. We can also execute a
subcomputation with a new masking state:\footnote{These functions take
  a \emph{higher-ranked} type. Removing the \texttt{forall} stuff, we
  have \texttt{(m a -> m a) -> m b}, which is a function which takes a
  function as an argument and returns a result. The \texttt{forall} is
  necessary because, without it, the concrete type that the variable
  \texttt{a} is unified with is fixed across \emph{all} usage sites,
  whereas with the \texttt{forall} it can be determined uniquely
  everywhere it is used.}

\begin{haskellcode}
mask :: MonadConc m => ((forall a. m a -> m a) -> m b) -> m b
uninterruptibleMask :: MonadConc m => ((forall a. m a -> m a) -> m b) -> m b
\end{haskellcode}

In both cases, the action evaluated is passed a function to reset the
masking state to the original one. A thread can be forked and given a
function to reset the masking state:

\begin{haskellcode}
forkWithUnmask :: MonadConc m => ((forall a. m a -> m a) -> m ())
  -> m (ThreadId m)
forkOnWithUnmask :: MonadConc m => Int -> ((forall a. m a -> m a) -> m ())
  -> m (ThreadId m)
\end{haskellcode}

We can also fork a thread and call a supplied function when the thread
is about to terminate, which is useful for informing the parent when a
child terminates, for example:

\begin{haskellcode}
forkFinally :: MonadConc m => m a -> (Either SomeException a -> m ())
  -> m (ThreadId m)
\end{haskellcode}

The \verb|SomeException| type is the top of the exception hierarchy,
and so can be used to catch all exceptions.

\subsubsection{Lifting Actions into \texttt{MonadConc}}
\label{sec:abstraction-typeclass-lifting}

If the programmer needs to make use of \verb|IO| actions, rather than
\verb|MonadConc| actions, then this can be achieved by adding a
\verb|MonadIO| context and using \verb|liftIO|. However, this can
easily compromise the results of testing, as the test runner cannot
peek inside \verb|IO| actions (that's why the typeclass exists in the
first place!). Thus, it is only safe if:

\begin{itemize}
\item \emph{The action is atomic and synchronised.}

  Otherwise the test framework will possibly miss schedules which lead
  to a bug.
\item \emph{The action is deterministic} (when executed as part of a
  computation with a deterministic schedule).

  Otherwise the fundamental assumption behind the testing methodology
  is false, and no guarantees about completeness can be made.
\item \emph{The action cannot block on the action of another thread.}

  Otherwise test execution may deadlock.
\end{itemize}
