There are currently 31 primitive actions used to construct the testing
instances of \verb|MonadConc|, one of which only arises when testing
under relaxed memory. These primitive actions contain a continuation,
allowing individual actions to be composed into larger execution
sequences. Each thread of execution consists of such a sequence,
terminated by the \verb|AStop| primitive, which has no continuation
and signals the termination of the thread.

Threads, \verb|CVar|s, \verb|CRef|s, and \verb|CTVar|s may be given
names, which are displayed in execution traces. If no name is given, a
unique numeric identifier is used instead.

\vspace{0.25cm}
\noindent \textsc{Threading} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{AFork}{name (unmask \arr action) (thread\_id \arr action)}{%
  Create a new thread from the first action, and continue executing
  the current thread with the second.}
\defineprimT{AMyTId}{(thread\_id \arr action)}{%
  Continue execution of the current thread by querying the thread
  identifier.}
\defineprimT{AYield}{action}{%
  Execute the given action, but also signify to the scheduler that it
  may be worth running a different thread now.}
\defineprimT{AStop}{}{%
  Terminate the current thread.}
\end{primtable}

\noindent \textsc{\texttt{CRef}s} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{ANewRef}{name a (cref a \arr action)}{%
  Construct a new \texttt{CRef} and give it to the continuation.}
\defineprimT{AReadRef}{(cref a) (a \arr action)}{%
  Read the currently visible value of a \texttt{CRef}.}
\defineprimT{AReadRefCas}{(cref a) (ticket a \arr action)}{%
  Produce a \texttt{Ticket} from the currently visible state of a
  \texttt{CRef}.}
\defineprimT{APeekTicket}{(ticket a) (a \arr action)}{%
  Get the value out of a \texttt{Ticket}.}
\defineprimT{AModRef}{(cref a) (a \arr (a, b)) (b \arr action)}{%
  Commit all pending writes and atomically modify the value within a
  \texttt{CRef}.}
\defineprimT{AModRefCas}{(cref a) (a \arr (a, b)) (b \arr action)}{%
  Commit all pending writes and atomically modify the value within a
  \texttt{CRef} using a compare-and-swap.}
\defineprimT{AWriteRef}{(cref a) a action}{%
  Update the value of a \texttt{CRef}. The updated value is visible to
  the current thread immediately.}
\defineprimT{ACasRef}{(cref a) (ticket a) a ((succeeded?, ticket a) \arr action)}{%
  Update the value of a \texttt{CRef} if it hasn't changed since the
  ticket was produced..}
\defineprimT{ACommit}{thread\_id cref\_id}{%
  Make the last write to the given \texttt{CRef} by that thread
  visible to all threads.}
\end{primtable}

\noindent \textsc{\texttt{CVar}s} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{ANewVar}{name (cvar a \arr action)}{%
  Construct a new \texttt{CVar} and give it to the continuation.}
\defineprimT{APutVar}{(cvar a) a action}{%
  Block until the \texttt{CVar} is empty and put a value into it.}
\defineprimT{ATryPutVar}{(cvar a) a (succeeded? \arr action)}{%
  Try to put a value into the \texttt{CVar} without blocking.}
\defineprimT{AReadVar}{(cvar a) (a \arr action)}{%
  Block until the \texttt{CVar} is full and read its value.}
\defineprimT{ATakeVar}{(cvar a) (a \arr action)}{%
  Block until the \texttt{CVar} is full and take its value.}
\defineprimT{ATryTakeVar}{(cvar a) (Maybe a \arr action)}{%
  Try to take the value from a \texttt{CVar} without blocking.}
\end{primtable}

\noindent \textsc{Exceptions} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{AThrow}{exception}{%
  Raises an exception in the current thread, terminating the current
  execution.}
\defineprimT{AThrowTo}{exception action}{%
  Raises an exception in the other thread, blocking if the other
  thread has exceptions masked.}
\defineprimT{ACatching}{(exception \arr handler) action continuation}{%
  Registers a new exception handler for the duration of the inner
  action.}
\defineprimT{APopCatching}{action}{%
  Remove the exception handler from the top of the stack.}
\defineprimT{AMasking}{masking\_state (unmask \arr action) continuation}{%
  Executes the inner action under a new masking state, and also gives
  it a function to reset the masking state.}
\defineprimT{AResetMask}{set? inner? masking\_state action}{%
  Sets the masking state.}
\end{primtable}

\noindent \textsc{Software Transactional Memory} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{AAtom}{transaction continuation}{%
  Execute an STM transaction atomically.}
\defineprimT{SNew}{name a (ctvar a \arr action)}{%
  Create a new \texttt{CTVar} containing the given value.}
\defineprimT{SRead}{(ctvar a) (a \arr action)}{%
  Read the current value of a \texttt{CTVar}.}
\defineprimT{SWrite}{(ctvar a) a action}{%
  Update a \texttt{CTVar}.}
\defineprimT{SThrow}{exception}{%
  Throw an exception, aborting the current execution flow.}
\defineprimT{SCatch}{(exception \arr handler) action continuation}{%
  Registers a new exception handler for the duration of the action.}
\defineprimT{SRetry}{}{%
  Abort the current transaction.}
\defineprimT{SOrElse}{transaction transaction continuation}{%
  Try executing the first transaction, if it fails, execute the
  second.}
\end{primtable}

\noindent \textsc{Testing Annotations} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{AKnowsAbout}{(Either cvar ctvar) action}{%
  Record that the thread has access to the given variable.}
\defineprimT{AForgets}{(Either cvar ctvar) action}{%
  Record that the thread no longer has access to the given variable.}
\defineprimT{AAllKnown}{action}{%
  Record that all variables the thread knows about have been
  reported.}
\end{primtable}

\noindent \textsc{Miscellaneous} \vspace{-0.5\parskip}

\begin{primtable}
\defineprimT{AReturn}{action}{%
  Execute the given action.}
\defineprimT{ALift}{monadic\_action}{%
  Execute an action from the underlying monad.}
\end{primtable}
