The \emph{Search Party}\footnote{\github{barrucadu}{search-party}}
library supports speculative parallelism in generate-and-test search
problems. It is motivated by the consideration that: if multiple
acceptable solutions exist, it may not matter which one is
returned. Initially, only single results could be returned, but
support for returning all results was later added, incorrectly,
introducing a bug.

The key piece of code causing the problem was this part of the worker
loop:

\begin{minted}{haskell}
case maybea of
  Just a -> do
    atomically $ do
      val <- tryTakeCTMVar res
      case val of
        Just (Just as) -> putCTMVar res $ Just (a:as)
        _ -> putCTMVar res $ Just [a]
    unless shortcircuit $
      process remaining res
  Nothing -> process remaining res
\end{minted}

Here \verb|maybea| is a value indicating whether the computation just
evaluated was successful. The intended behaviour is that, if a
computation is successful, its result is added to the list in the
\verb|res| \verb|CTMVar|. This \verb|CTMVar| is exposed indirectly to
the user of the library, as it is blocked upon when the final result
of the search is requested.

There are some small tests in \emph{Search Party}, verifying that
deadlocks and exceptions don't arise, and that results are as
expected. Upon introducing this new functionality, tests began to fail
with differing result lists returned for different schedules,
prompting the test:\footnote{The \texttt{representative} function
  picks only one trace for each unique result.}

\begin{minted}{haskell}
checkResultLists :: Eq a => Predicate [a]
checkResultLists = representative (alwaysTrue2 check) where
  check (Right (Just as)) (Right (Just bs)) = as `elem` permutations bs
  check a b = a == b
\end{minted}

Given this predicate, we can very clearly see the problem:

\begin{verbatim}
> dejafu (runFind $ [0..2] @! const True) ("Result Lists", checkResultLists)

[fail] Result Lists (checked: 145)
        Just [0] S0-----S1---------S3-------S0-------
        Just [1] S0-----S1---------S3---P2-------S0-------
        Just [1,0] S0-----S1---------S3-------S2-------S0-------
        Just [0,1] S0-----S1---------S3---P2-------S3----S0-------
False
\end{verbatim}

The problem was a lack of any indication that a list-producing
computation had finished. As results were written directly to the
\verb|CTMVar|, partial result lists could be read depending on how the
worker threads and the main thread were interleaved.

In this case, fixing the failure did not require any interactive
debugging. Only one place had been modified in introducing the new
functionality, and the bug was found by re-reading the code with the
possibility of error in mind. However, the ability to produce a test
case which reliably reproduces the problem gives confidence that it
will not be accidentally reintroduced.
